Co-methylation

### 2014-10-02 Rmarkdown/comethylation.Rmd
Reviewed Lacey (2013), Hebestreit (2013) and Liu (2014).

### Correlations of aggregate measurements of methylation

To date, most analyses of co-methylation have been based on aggregate measurements of methylation, such as $\beta$-values \cite{Eckhardt:2006gh, Cokus:2008fc, Li:2010fb, Lyko:2010dr, Hebestreit:2013ko, Lacey:2013iy, Liu:dy}. This is true of studies using both microarray data, where only these type of measurements are available, and sequencing data, where base-resolution measurements are also available. Very often the exact analysis is not well-described but it can generally be thought of as studying the autocorrelation of the $\beta$-values along the genome.

@@

\cite{Eckhardt:2006gh} is an important paper in the field and has been cited as evidence that "methylation levels are strongly spatially correlated" \cite{Hebestreit:2013ko} and that "proximal CpGs [have] similar methylation levels" \cite{Hansen:2011gu}. \cite{Eckhardt:2006gh} performed  PCR-amplicon bisulfite-sequencing of 2,524 amplicons on human chromosomes 6, 20 and 22.  PCR-amplicon bisulfite-sequencing is a labour-intensive assay and so it was a considerable effort to produce these data. Nonetheless, it is important to bear in mind that these data are nowhere near as comprehensive as data generated using modern-day whole-genome bisulfite-sequencing assays. Samples were provided by male and female donors across a wide variety of tissue types and ages, and were were generated to begin a catalogue of CpG methylation in various tissue types.

@@

#### Other co-methylation analyses using aggregate measurements

* \cite{Hebestreit:2013ko, Lacey:2013iy, Liu:dy}

\paragraph{Previous within-fragment co-methylation analyses}
\label{sec:previous_within-fragment_analyses}

* More recently, within-fragment analyses. Describe concept (move from next section to this section).
* Cokus

#### Other definitions

@@

@@

\subsection{Within-fragment co-methylation}\label{subsec:wf_cometh}

@@

\subsubsection{Using pairs of methylation loci}\label{subsec:pairs}

@@

#### Comparison of LOR to correlations

@@

#### Constructing pairs of methylation loci

@@

#### Visualisations

@@

### Using higher-order $m$-tuples of methylation loci

@@

#### An EM algorithm to estimate co-methylation

@@

## Correlations of $\beta$-values

>>>>

\subsection{Correlations of aggregate measurements of methylation}
\label{sec:previous_cor_of_agg_methylation}

To date, most analyses of co-methylation have been based on analysing aggregate measurements of methylation, such as $\beta$-values, within individual samples \cite{Eckhardt:2006gh, Cokus:2008fc, Li:2010fb, Lyko:2010dr, Lacey:2013iy}. These analyses of within-sample correlations of aggregate methylation levels have an implicit assumption of stationarity along the genome (__TODO: Discuss with Terry__).

Here I review this previous work. I have been forced to make my "best guess" about some of the methods since they are often not sufficiently described in the original publication.

@@

\cite{Eckhardt:2006gh} is an important paper and one of the first to apply bisulfite-sequencing to a large number of regions and samples. Of particular relevance to a discussion of co-methylation, \cite{Eckhardt:2006gh} has been cited as evidence that "methylation levels are strongly spatially correlated" \cite{Hebestreit:2013ko} and that "proximal CpGs [have] similar methylation levels" \cite{Hansen:2011gu}. \cite{Eckhardt:2006gh} performed  PCR-amplicon bisulfite-sequencing of 2,524 amplicons on human chromosomes 6, 20 and 22.  PCR-amplicon bisulfite-sequencing is a labour-intensive assay and so it was a considerable effort to produce these data. Nonetheless, it is important to bear in mind that these data are nowhere near as comprehensive as data generated using modern-day whole-genome bisulfite-sequencing assays. Samples were provided by male and female donors across a wide variety of tissue types and ages, and were were generated to begin a catalogue of CpG methylation in various tissue types.

@@

#### \cite{Lacey:2013iy}

\cite{Lacey:2013iy} fit a Gaussian variogram model to the $\beta$-values at CpGs obtained via reduced representation bisulfite-sequencing (RRBS) of chromosome 11 from a normal myotube cell line (MTCTL2). The empirical and fitted variograms flatten out for CpGs separated by around $3000$ bp (Figure 1d) and so they conclude that the methylation level of CpGs "show a strong correlation for sites in close proximity, decaying to near independence at distances beyond 3000 bp".

Caution is warranted in extrapolating these results to statements about genome-wide co-methylation. Firstly, this analysis uses only approximately $60,000$ CpGs[^ncpgs], far fewer than the $25$ million CpGs (__TODO: Check number__) in the haploid human genome. Secondly, these data are from a single sample from a single tissue, which cannot capture the sample-to-sample nor tissue-to-tissue variability in co-methylation. Finally, RRBS targets CpG dense regions of the genome with low levels of methylation and so these data do not capture co-methylation outside of these regions, where the majority of the CpGs are found and which are highly methylated.

[^ncpgs]: The data for MTCTL2 are not publicly available so I estimated the number of sequenced CpGs on chromosome 11 from three comparable RRBS libraries (__TODO: Cite and link to appendix__).

\subsection{Within-fragment co-methylation analyses}
\label{sec:previous_within-fragment_analyses}

* More recently, within-fragment analyses. Describe concept (move from next section to this section).
* Cokus
* \cite{Lacey:2009by}
* \cite{Paliwal:2013fk}

\subsection{Other related work}
\label{sec:previous_other_related_work}

Other definitions of co-methylation and related work:

@@

#### \cite{Hebestreit:2013ko}

The `BiSeq` R/Bioconductor software uses a semivariogram to estimate the spatial correlation between test statistics of differential methylation at two methylation loci \cite{Hebestreit:2013ko}. This is not the same as looking for correlations between the $\beta$-values but rather looking for correlations between the test statistics, which themselves are based on _differences_ of $\beta$-values.

#### \cite{Liu:dy}

The focus of this study on identify clusters of correlated CpGs and relating methylation changes at these clusters to SNP genotypes. The authors compute the correlation of pairs of $\beta$-values across samples, which is analogous to estimating the linkage disequilibrium between a pair of single nucleotide polymorphisms (SNPs) across a set of samples. This is different to computing autocorrelations of $\beta$-values, which are computed within, rather than between, samples. An autocorrelation analysis implicitly assumes stationarity along the genome whereas this analysis assumes stationarity across samples (__TODO: Discuss with Terry__).

The discovery cohort contains 247 human samples typed on the Illumina HumanMethylation450 array. After normalisation and other pre-processing steps, they identify the top-$25\%$ most variably methylated CpG sites (vCpGs) in the 247 samples. They then compute Pearson correlations of methylation levels for all pairs of vCpGs across the samples.

From this they conclude that "DNA methylation is correlated over regions with a median length of 274 bp from our data set". However, they also note that "methylation codependency (co-methylation) itself varies across the genome", as is to be expected.

@@

\section{Correlations of $\beta$-values}\label{subsec:beta_cor}

\section{Within-fragment co-methylation}\label{subsec:wf_cometh}

@@

\subsection{Using pairs of methylation loci}\label{subsec:pairs}

@@

### Comparison of LOR to correlations

@@

### Constructing pairs of methylation loci

@@

### Visualisations

@@

## Using higher-order $m$-tuples of methylation loci

@@

### An EM algorithm to estimate co-methylation

@@
